---
title: "Whether or not to normalize time by D_I"
author: "Watal M. Iwasaki"
date: "2014-09-18"
output:
  html_document:
    toc: true
---

Codes
================================================================================

```{r, echo=FALSE, eval=FALSE}
.outdir = '~/edal/html/results'
rmarkdown::render('~/edal/normalizing_time.Rmd', output_dir=.outdir)
```

```{r library}
library(plyr)
library(dplyr)
library(stringr)
library(tidyr)
library(pipeR)
library(ggplot2)
library(gridExtra)
library(doMC)
doMC::registerDoMC(min(parallel::detectCores(), 12))
```

```{r preparation}
.read_conf = function(.rundir) {
    .conf_file = list.files(.rundir, '\\.conf$', full.names=TRUE)
    wtl::read.conf(.conf_file) %>>%
        dplyr::select(-help, -test, -verbose, -seed, -ppn)
}

.traits = list(
    toepad_P=expression(paste('Toepad size ', italic(x[0]))),
    limb_P=expression(paste('Limb length ', italic(x[1]))),
    height_pref_P=expression(paste('Height pref. ', italic(y[0]))),
    diameter_pref_P=expression(paste('Diameter pref. ', italic(y[1]))),
    male_P=expression(paste('Male trait ', italic(m))),
    female_P=expression(paste('Female trait ', italic(f))),
    choosiness_P=expression(paste('Choosiness ', italic(c))),
    neutral_P='Neutral'
)

.tr = .traits
.axes = matrix(names(.traits), ncol=2, byrow=TRUE)

theme_tile =
    theme(panel.background=element_rect(fill='white', colour='black'))+
    theme(panel.grid=element_blank())+
    theme(axis.ticks=element_blank())+
    theme(axis.text=element_blank())+
    theme(legend.position='none')
```

```{r plot_each_result, fig.width=9, fig.height=9, echo=FALSE, eval=FALSE}
.labeller = function(varname, value) {
    .traits[as.character(value)]
}

.plot_a_result = function(.data) {.data %>>%
    select(ends_with('P')) %>>%
    gather_('xkey', 'xval', .axes[,1]) %>>%
    gather_('ykey', 'yval', .axes[,2]) %>>%
    filter(charmatch(xkey, .axes[,1]) == charmatch(ykey, .axes[,2])) %>>%
    group_by(xkey, ykey, xval, yval) %>>% tally() %>>%
    ggplot(aes(xval, yval))+
        geom_tile(aes(fill=n))+
        scale_fill_gradient(low='#EEEEEE', high='darkcyan')+
        facet_grid(~ xkey + ykey, labeller=.labeller)+
        coord_cartesian(c(0, 1), c(0, 1))+
        theme_tile+
        theme(axis.title=element_blank())
}
.aggr = function(x) {
    .label = unique(x$label)
    .gpl = plyr::llply(x$.id, function(.rundir) {
        cat(.rundir, '\n')
        .raw = read.csv(file.path(.rundir, 'population.csv.gz'))
        .plot_a_result(.raw)
    }, .parallel=TRUE)
    .grob = do.call(gridExtra::arrangeGrob, c(.gpl, list(nrow=4, ncol=1)))
    .grob
}

.working_dir = '~/SpiderOak Hive/anole20140130'
.indirs = list.files(.working_dir, full.names=TRUE)
.indirs = .indirs[isdir(.indirs)]
names(.indirs) = .indirs
.conf = plyr::ldply(.indirs, .read_conf)

.tbl = .conf %>>%
    filter(str_detect(label, '^c')) %>>%
    group_by(label) %>>%
    do(plot=.aggr(.))
.tbl$plot
```

```{r plot_comparison}
.data_dirs = c('~/SpiderOak Hive/anole20140130', '~/working/anolis20140903')
.indirs = list.files(.data_dirs, full.names=TRUE)
.indirs = .indirs[isdir(.indirs)]
names(.indirs) = .indirs
.conf = plyr::ldply(.indirs, .read_conf)

.tbl = .conf %>>%
    mutate(normalized=str_detect(.id, 'anole20140130')) %>>%
    group_by(normalized, label) %>>%
    do(cbind(., replicate=seq_len(nrow(.)))) %>>%
    rowwise() %>>%
    do(cbind(.,
        file.path(.$.id, 'population.csv.gz') %>>%
        read.csv(stringsAsFactors=FALSE) %>>%
        select(-ends_with('L'), -ends_with('R')), stringsAsFactors=FALSE)) %>>% ungroup()

.done = .tbl %>>%
    group_by(.id, normalized, label, replicate, height_pref_P, diameter_pref_P) %>>%
    tally() %>>%
    group_by(param=substr(label, 1, 1)) %>>%
    do(plot={. %>>%
        ggplot(aes(height_pref_P, diameter_pref_P))+
        geom_tile(aes(fill=n))+
        scale_fill_gradient(low='#DDDDDD', high='#FF3300')+
        facet_grid(label ~ normalized + replicate)+
        labs(x=.tr[['height_pref_P']], y=.tr[['diameter_pref_P']])+
        #, labeller=function(varname, value) {str_extract(basename(value), '^[^_]+')})+
        coord_cartesian(c(0, 1), c(0, 1))+
        theme_tile
    })
```

Results
================================================================================

Each heatmap denotes the frequency disturibution of habitat preferences in a population.
Four replicates were performed for each parameter set (i.e. each row).
The results with **normalized time** described in anolis_v3.pdf are on the right (`TRUE`) and **unnormalized** on the left (`FALSE`).

In unnormalized cases, on the left, the habitat preferences of individuals are distributed near the peak of resource abundance and seem to be properly subject to selection.
The variation within each population decreased with increasing coefficient for habitat preferences, $h_0$ and $h_1$ (from top to bottom).

```{r habitat_pref, echo=FALSE, fig.width=8, fig.height=5}
subset(.done, param=='p')$plot[[1]]
```

Other parameters did not affect the evolution of habitat preferences so much.
Competition coefficients, $c_0$ and $c_1$, were altered:

```{r competition, echo=FALSE, fig.width=8, fig.height=5}
subset(.done, param=='c')$plot[[1]]
```

Selection coefficients for morphology, $s_0$ and $s_1$, were altered:

```{r morph_selec, echo=FALSE, fig.width=8, fig.height=5}
subset(.done, param=='s')$plot[[1]]
```

Scaling parameter $\sigma_a$ for female mating preferences was altered:

```{r mating, echo=FALSE, fig.width=8, fig.height=5}
subset(.done, param=='f')$plot[[1]]
```
